A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-11-09 jour/jour.tex
Starting work on extending sec 4.

\section{Implementation as a Theory Solver in CVC4}
\label{sec:implementation-as-a-theory-solver-in-cvc4}

@@

%%% @ANDY: I'm trying without \paragraph's, to save space but also to make the
%%% section look more like a whole than a random list of stuff. Revert if you
%%% disagree.

The decision procedure is implemented as a \relax{theory solver} of CVC4, that is, a
specialized solver for determining the satisfiability of conjunctions of
literals for its theory.
Given a theory $T = T_1 \mathrel\cup \cdots \mathrel\cup T_n$ and a set of
input clauses $F$ in conjunctive normal form, the CDCL($T$) procedure
incrementally builds partial assignments of
truth values to the atoms of $F$ such that no clause in $F$ is falsified. We can regard such a
partial assignment as a set $M$ of true literals.
% where $a \in M$ for all atoms $a$ that
%it assigned the value $\top$ and $\neg\: a \in M$ for all atoms that it assigns
%the value $\bot$.
%at a high level,
By a variant of the Nelson--Oppen method \cite{jovanovic2011sharing,nelson-oppen-1979},
each $T_i$-solver %for $i = 1, \ldots, l$  n?
takes as input the union $M_i$
of (1)~the purified form of $T_i$-literals occurring in $M$, where fresh
variables replace terms containing symbols not belonging to~$T_i$;
(2)~additional (dis)equalities between variables of types not belonging to
$T_i$. Each $T_i$-solver
either reports that a subset $C$ of $M_i$ is $T_i$-unsatisfiable, in which case
$\neg\: C$ is added to $F\!$, adds a clause to $F\!$, or does nothing.
When $M$ is a complete assignment for $F\!$, a theory solver can choose to do
nothing only if $M_i$ is indeed $T_i$-satisfiable.

@@

While the decision procedure is restricted to universal conjectures, users often
want to solve problems that feature universal axioms and existential
conjectures.
%%% @ANDY: not sure how to phrase this now
%where checking satisfiability requires reasoning about ground literals in the
%presence of universally quantified formulas.
Many SMT solvers, including CVC4, can reason about quantified formulas using
incomplete instantiation-based
techniques~\cite{MouraBjoerner07,ReynoldsTinelliMoura14}.
These techniques extend naturally to %quantified
formulas involving datatypes and codatatypes.

@@

>>>>

\section{Implementation in CVC4}
\label{sec:implementation-as-a-theory-solver-in-cvc4}

@@

%%% @ANDY: I'm trying without \paragraph's, to save space but also to make the
%%% section look more like a whole than a random list of stuff. Revert if you
%%% disagree.
\subsection{Theory Solver for $\thD$}
The decision procedure is implemented as a \relax{theory solver} of CVC4, that is, a
specialized solver for determining the satisfiability of conjunctions of
literals for its theory.
Given a theory $T = T_1 \mathrel\cup \cdots \mathrel\cup T_n$ and a set of
input clauses $F$ in conjunctive normal form, the CDCL($T$) procedure
incrementally builds partial assignments of
truth values to the atoms of $F$ such that no clause in $F$ is falsified. We can regard such a
partial assignment as a set $M$ of true literals.
% where $a \in M$ for all atoms $a$ that
%it assigned the value $\top$ and $\neg\: a \in M$ for all atoms that it assigns
%the value $\bot$.
%at a high level,
By a variant of the Nelson--Oppen method \cite{jovanovic2011sharing,nelson-oppen-1979},
each $T_i$-solver %for $i = 1, \ldots, l$  n?
takes as input the union $M_i$
of (1)~the purified form of $T_i$-literals occurring in $M$, where fresh
variables replace terms containing symbols not belonging to~$T_i$;
(2)~additional (dis)equalities between variables of types not belonging to
$T_i$. Each $T_i$-solver
either reports that a subset $C$ of $M_i$ is $T_i$-unsatisfiable, in which case
$\neg\: C$ is added to $F\!$, adds a clause to $F\!$, or does nothing.
When $M$ is a complete assignment for $F\!$, a theory solver can choose to do
nothing only if $M_i$ is indeed $T_i$-satisfiable.

@@

@@

\begin{figure}[t!]
\normalsize
\centering
\begin{enumerate}
\item Let $\mathcal{U}$ be a partition of the equivalence classes of $\Ec$ such that $\ec{x} =_{\mathcal{U}} \ec{y}$ if and only if:
\begin{itemize}
\item $\const{C}( x_1, \ldots, x_k ) \in \ec{x}$ and $\const{C}( y_1, \ldots, y_k ) \in \ec{y}$.
\end{itemize}
\item Repeat until $\mathcal{U}$ is unchanged
\begin{enumerate}
\item $\mathcal{U}' := \emptyset$
\item for each $U_i \in \mathcal{U}$:
\begin{itemize}
\item[\ ] Let $\mathcal{U}_i$ be a partition of $U_i$ such that $\ec{x} =_{\mathcal{U}_i} \ec{y}$ for distinct $\ec{x}, \ec{y}$ if and only if:
\begin{itemize}
\item $\const{C}( x_1, \ldots, x_k ) \in \ec{x}$ and $\const{C}( y_1, \ldots, y_k ) \in \ec{y}$, and
\item $\ec{x_j} =_{\mathcal{U}} \ec{y_j}$ for each $j = 1, \ldots, k$.
\end{itemize}
\item[\ ] $\mathcal{U}' := \mathcal{U}' \cup \mathcal{U}_i$.
\end{itemize}
\item $\mathcal{U} := \mathcal{U}'$
\end{enumerate}
\item If distinct $t\typ{\delta}, s\typ{\delta} \in U \in \mathcal{U}$ where $\delta \in \Codata$, apply \rn{Unique} to $t,s$.
\end{enumerate}
\vspace*{-2pt} %% TYPESETTING HACK
\caption{\,Algorithm for checking Uniqueness.
}
\label{fig:cc-rules}
\end{figure}

\subsection{Model Construction}

[TODO]

\subsection{Extension to Quantified Formulas}
While the decision procedure is restricted to universal conjectures, users often
want to solve problems that feature universal axioms and existential
conjectures.
%%% @ANDY: not sure how to phrase this now
%where checking satisfiability requires reasoning about ground literals in the
%presence of universally quantified formulas.
Many SMT solvers, including CVC4, can reason about quantified formulas using incomplete 
instantiation-based 
techniques~\cite{MouraBjoerner07,ReynoldsTinelliMoura14}.
These techniques extend naturally to %quantified
formulas involving datatypes and codatatypes.
